Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-20 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, i.e.~dissolution, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Some examples in the figure include the rough surface of limestone pavement, a sinkhole relic in Belize, and an erosion-curved conduit. The final image shows the small-scale structure of soil which exhibits polydispersity and anisotropy resulting from a combination of erosive, compaction, and transport actions.

@@

{\bf An ultimate goal of this proposal is to simulate $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with high-order accuracy and the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

@@

In addition to the work of the PIs, dynamic erosion of a porous media is a highly active subject as laboratory experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, {\bf we propose performing a set of experiments conducted at the nearby Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).}

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing represent rapid changes.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.  
\todo[inline]{The last paragraph of this sections seems out of place.  Maybe it just needs to be reversed with the one before it?}
%----------------------------------------------------------------------%
\subsection{Theoretical Foundation and Governing Equations}

@@

>>>>

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of porous-medium constituents, enabling smaller fragments to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of fragments or grains. For instance, erosion can carve grains  into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, i.e.~dissolution, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Some examples in the figure include the rough surface of limestone pavement, a sinkhole relic in Belize, and an erosion-curved conduit. The final image shows the small-scale structure of soil which exhibits polydispersity and anisotropy resulting from a combination of erosive, compaction, and transport actions.

@@

{\bf An ultimate goal of this proposal is to simulate $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with high-order accuracy and the computational efficiency required to run large ensembles for statistical analysis.} 
This topic of the dynamic erosion of a porous medium is one that has been receiving significant recent interest. Laboratory experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison.

@@

A major step towards the goal of developing high-fidelity simulations of dynamic porous-medium erosion is illustrated in Fig.~\ref{80circ8vel}. This figure shows simulated erosion of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing represent rapid changes.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse. Furthermore, we propose performing a set of experiments conducted at GFDI that would help to guide, calibrate, and validate the computations; See Section~\ref{sec:experiments} and the letter of collaboration for further details.

\todo[inline]{Okay, I did some rearranging. See if you like it better.}

%----------------------------------------------------------------------%
\subsection{Theoretical Foundation and Governing Equations}

@@

\item The coarse-grained models will harness recent advances in deep neural networks \cite{ling2016reynolds, kutz2017deep, ling2016machine, pelt2018mixed} to parameterize the effects of mechanical and chemical erosion within a multiphase\cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase}, or similar macroscopic, framework.
